Finiteness in Cubical Type Theory

### 2020-01-23 paper.tex
more

\begin{romdefinition}[Fibers] \label{fibers}
  A fiber is defined over some function \(f : A \rightarrow B\).

@@

Split enumerability implies decidable equality on the underlying type.
To prove this, we will make use of the following lemma, proven in the
formalisation:
\begin{romlemma} \label{discrete-surj}
  \begin{equation}
    \frac{
        A \twoheadrightarrow! \; B \; \; \; \text{Discrete}(A)
      }{
       \text{Discrete}(B) 
      }
  \end{equation}
\end{romlemma}
\begin{romlemma} \label{split-enum-discrete}
  Every split enumerable type is discrete.
\end{romlemma}
\begin{proof}
  Let \(A\) be a split enumerable type.
  By lemma~\ref{split-enum-is-split-surj}, there is a surjection from
  \(\mathbf{Fin}(n)\) for some \(n\).
  Also, we know that \(\mathbf{Fin}(n)\) is discrete (proven in our
  formalisation).
  Therefore, by lemma~\ref{discrete-surj}, \(A\) is discrete.
  \qed
\end{proof}
\subsection{Manifest Bishop Finiteness}
\begin{romdefinition}[Manifest Bishop Finiteness]
  \begin{equation} \label{bish-def}
    \mathcal{B}(A) \coloneqq \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \inunique xs
  \end{equation}
\end{romdefinition}
The only difference between manifest Bishop finiteness and split enumerability
is the membership term: here we require unique membership (\(\inunique\)),
rather than simple membership (\(\in\)).
\begin{romdefinition}[Unique Membership] \label{uniq-memb-def}
  \begin{equation}
    x \inunique \mathit{xs} \coloneqq \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{romdefinition}

@@

To derive \(\mathcal{B}\) from \(\mathcal{E}!\) takes significantly more work.
The ``unique membership'' condition in \(\mathcal{B}\) means that we are not
permitted duplicates in the support list.
The first step in the proof, then, is to filter those duplicates out from the
support list of the \(\mathcal{E}!\) proof: we can do this using the decidable
equality provided by \(\mathcal{E}!\) (lemma~\ref{split-enum-discrete}).
From there, all we need to show is that the membership proof carries over
appropriately.
\begin{romlemma} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{romlemma}
\subsection{Manifest Bishop Finiteness and Manifest Enumerability}
\begin{romlemma}
  Any manifestly Bishop finite type is manifestly enumerable.
\end{romlemma}
\begin{romlemma}
  A manifestly enumerable type with decidable equality is manifest Bishop
  finite.
\end{romlemma}
\subsection{Manifest Bishop Finiteness and Cardinal Finiteness}
\begin{romlemma}
  Any manifest Bishop finite type is cardinal finite.
\end{romlemma}
\begin{romtheorem}
  Any cardinal finite type with a total order is Bishop finite.
\end{romtheorem}
\subsection{Manifest Enumerability and Kuratowski Finiteness}
\begin{romlemma}
  Any manifest enumerable type is Kuratowski finite.
\end{romlemma}
\subsection{Cardinal Finiteness and Kuratowski Finiteness}
\begin{romtheorem} \label{kuratowski-cardinal}
  \begin{equation}
    \mathcal{K}^f(A) \times \text{Discrete}(A) \simeq \mathcal{C}(A)
  \end{equation}
\end{romtheorem}
\section{Topos}
Our main result in this section is the formal proof that decidable Kuratowski
finite types form a \(Pi\)-pretopos.
As we saw in theorem~\ref{kuratowski-cardinal}, decidable Kuratowski finite
types are equivalent to cardinal finite types, so we will work with the latter
from now on.
Our first task is to show that cardinal finite types are closed under several
operations.
\subsection{Closure}
For the first two closure proofs, we only consider manifest Bishop finiteness:
as it is the strongest of the finiteness predicates, we can derive the other
closure proofs from it.
\begin{romlemma}
  \(\bot\) is manifest Bishop finite.
\end{romlemma}
\begin{proof}
  From lemma~\ref{bishop-equiv} we can prove the goal by constructing a term of
  the type:
  \begin{equation}
    \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \simeq \bot)
  \end{equation}
  To do so, we take \(n\) to be 0, and by definition~\ref{Fin},
  \(\mathbf{Fin}(0) = \bot\).
  The equivalence is simply the identity equivalence.
\end{proof}
\begin{romlemma}
  \(\top\) is manifest Bishop finite.
\end{romlemma}
\begin{proof}
  Again, we need only provide a suitable cardinality (in this case 1), and an
  equivalence with \(\mathbf{Fin}(1)\) to prove our goal.
  \begin{align*}
    \top &\simeq \mathbf{Fin}(1) \\
          &\simeq \top \uplus \mathbf{Fin}(0) \\
          &\simeq \top \uplus \bot \\
          &\simeq \top
  \end{align*}
  \qed
\end{proof}
\begin{romlemma}
  \(\mathbf{Bool}\) is manifest Bishop finite.
\end{romlemma}
\begin{proof}
  We prove this by showing an isomorphism with \(\mathbf{Fin}(2)\).
  \begin{align*}
    \text{false} & \leftrightarrow 0 \\
    \text{true} & \leftrightarrow 1
  \end{align*}
  \qed
\end{proof}

@@

  We then prove an isomorphism between this representation and \(\Pi\)-types.
  \begin{equation}
    \mathbf{Tuple}(T, n) \iff \Pi {(x : \mathbf{Fin}(n))} , T(x)
  \end{equation}
  This allows us to transport our proof of finiteness on tuples to one on
  functions from \(\mathbf{Fin}\) (our second transport operation), proving our
  goal.
\end{proof}
\subsection{The Category of Finite Sets}
\subsection{The \(\Pi\)-pretopos of Finite Sets}
\section{Countably Infinite Types}
\begin{itemize}
  \item Definition
  \item \(\Sigma\)
  \item Kleene star
  \item Truncated
  \item Topos
\end{itemize}
\section{Search}
\begin{itemize}
  \item Omniscience
  \item Search
  \item Curry / uncurry
  \item Synth function
\end{itemize}

>>>>

\begin{romdefinition}[Fibers] \label{fibres}
  A fibre is defined over some function \(f : A \rightarrow B\).

@@

Split enumerability implies decidable equality on the underlying type.
To prove this, we will make use of the following lemma, proven in the
formalisation:
\begin{romlemma} \label{discrete-surj}
  \begin{equation}
    \frac{
        A \twoheadrightarrow! \; B \; \; \; \text{Discrete}(A)
      }{
       \text{Discrete}(B) 
      }
  \end{equation}
\end{romlemma}
\begin{romlemma} \label{split-enum-discrete}
  Every split enumerable type is discrete.
\end{romlemma}
\begin{proof}
  Let \(A\) be a split enumerable type.
  By lemma~\ref{split-enum-is-split-surj}, there is a surjection from
  \(\mathbf{Fin}(n)\) for some \(n\).
  Also, we know that \(\mathbf{Fin}(n)\) is discrete (proven in our
  formalisation).
  Therefore, by lemma~\ref{discrete-surj}, \(A\) is discrete.
  \qed
\end{proof}
\subsection{Manifest Bishop Finiteness}
\todo[noline]{Add explanation for word ``manifest''}
\begin{romdefinition}[Manifest Bishop Finiteness]
  \begin{equation} \label{bish-def}
    \mathcal{B}(A) \coloneqq \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \inunique xs
  \end{equation}
\end{romdefinition}
The only difference between manifest Bishop finiteness and split enumerability
is the membership term: here we require unique membership (\(\inunique\)),
rather than simple membership (\(\in\)).
\begin{romdefinition}[Unique Membership] \label{uniq-memb-def}
  \begin{equation}
    x \inunique \mathit{xs} \coloneqq \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{romdefinition}

@@

To derive \(\mathcal{B}\) from \(\mathcal{E}!\) takes significantly more work.
The ``unique membership'' condition in \(\mathcal{B}\) means that we are not
permitted duplicates in the support list.
The first step in the proof, then, is to filter those duplicates out from the
support list of the \(\mathcal{E}!\) proof: we can do this using the decidable
equality provided by \(\mathcal{E}!\) (lemma~\ref{split-enum-discrete}).
From there, all we need to show is that the membership proof carries over
appropriately.
\begin{romlemma} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{romlemma}
\subsection{Manifest Bishop Finiteness and Manifest Enumerability}
\begin{romlemma}
  Any manifestly Bishop finite type is manifestly enumerable.
\end{romlemma}
\begin{romlemma}
  A manifestly enumerable type with decidable equality is manifest Bishop
  finite.
\end{romlemma}
\subsection{Manifest Bishop Finiteness and Cardinal Finiteness}
\begin{romlemma}
  Any manifest Bishop finite type is cardinal finite.
\end{romlemma}
\begin{romtheorem}
  Any cardinal finite type with a total order is Bishop finite.
\end{romtheorem}
\section{Topos}
Our main result in this section is the formal proof that decidable Kuratowski
finite types form a \(Pi\)-pretopos.
As we saw in theorem~\ref{kuratowski-cardinal}, decidable Kuratowski finite
types are equivalent to cardinal finite types, so we will work with the latter
from now on.
Our first task is to show that cardinal finite types are closed under several
operations.
\subsection{Closure}
For the first two closure proofs, we only consider manifest Bishop finiteness:
as it is the strongest of the finiteness predicates, we can derive the other
closure proofs from it.
\begin{romlemma}
  \(\bot\) is manifest Bishop finite.
\end{romlemma}
\begin{proof}
  From lemma~\ref{bishop-equiv} we can prove the goal by constructing a term of
  the type:
  \begin{equation}
    \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \simeq \bot)
  \end{equation}
  To do so, we take \(n\) to be 0, and by definition~\ref{Fin},
  \(\mathbf{Fin}(0) = \bot\).
  The equivalence is simply the identity equivalence.
\end{proof}
\begin{romlemma}
  \(\top\) is manifest Bishop finite.
\end{romlemma}
\begin{proof}
  Again, we need only provide a suitable cardinality (in this case 1), and an
  equivalence with \(\mathbf{Fin}(1)\) to prove our goal.
  \begin{align*}
    \top &\simeq \mathbf{Fin}(1) \\
          &\simeq \top \uplus \mathbf{Fin}(0) \\
          &\simeq \top \uplus \bot \\
          &\simeq \top
  \end{align*}
  \qed
\end{proof}
\begin{romlemma}
  \(\mathbf{Bool}\) is manifest Bishop finite.
\end{romlemma}

@@

  We then prove an isomorphism between this representation and \(\Pi\)-types.
  \begin{equation}
    \mathbf{Tuple}(T, n) \iff \Pi {(x : \mathbf{Fin}(n))} , T(x)
  \end{equation}
  This allows us to transport our proof of finiteness on tuples to one on
  functions from \(\mathbf{Fin}\) (our second transport operation), proving our
  goal.
\end{proof}
\subsection{The Category of Finite Sets}
HoTT and CuTT seem to be especially suitable settings for formalisations of
category theory.
The univalence axiom in particular allows us to treat categorical isomorphisms
as the important, useful objects that they are, upgrading them to full
equalities.

We follow \cite[chapter 9]{hottbook} in its treatment of
categories in HoTT, and in its proof that sets do indeed form a category.
We will first briefly go through the construction of the category
\(\mathit{Set}\), as it differs slightly from the usual method in type theory.

First, the type of objects and arrows:
\begin{alignat}{3}
  &\text{Obj}_\mathit{Set}      &&\coloneqq \Sigma(x : \mathbf{Type}) , \text{isProp}(x) \\
  &\text{Hom}_\mathit{Set}(x , y) &&\coloneqq  \text{fst}(x) \rightarrow \text{fst}(y)
\end{alignat}
As is plain in the type of objects, we have already departed slightly from the
simpler \(\text{Obj}_\mathit{Set} \coloneqq \mathbf{Type}\) way of doing things:
of course we have to, as HoTT allows non-set types.
Furthermore, after proving the usual associativity and identity laws for
composition (which are definitionally true in this case), we must further show
\(\text{isSet}(\text{Hom}_\mathit{Set}(x,y))\); even then we only have a
precategory.

To show that \(\mathit{Set}\) is a category, we must show that categorical
isomorphisms are equivalent to equivalences.
In a sense, we must give a univalence rule for the category we are working in.

We have provided formal proofs that \(\mathit{Set}\) does indeed form a
category, and the following:
\begin{romtheorem}[The Category of Finite Sets]
  Finite sets form a category in HoTT when defined like so:
  \begin{equation}
    \begin{aligned}
      &\text{Obj}_\mathit{FinSet}      &&\coloneqq \Sigma(x : \mathbf{Type}) , \mathcal{C}(x) \\
      &\text{Hom}_\mathit{FinSet}(x , y) &&\coloneqq  \text{fst}(x) \rightarrow \text{fst}(y)
    \end{aligned}
  \end{equation}
\end{romtheorem}
\subsection{The \(\Pi\)-pretopos of Finite Sets}
For this proof, we follow again the proof that \(\mathit{Set}\) forms a \(\Pi
W\)-pretopos from \cite[chapter 10]{hottbook} and
\cite{rijkeSetsHomotopyType2015}.
\section{Countably Infinite Types}
In the previous sections we saw different flavours of finiteness which were
really just different flavours of relations to \(\mathbf{Fin}\).
In this section we will see that we can construct a similar classification of
relations to \(\mathbb{N}\), in the form of the countably infinite types.
\subsection{Split Countable Types}
Our first foray into the world of countable types will be a straightforward
analogue to the split enumerable types.
We need change only one element: instead of a support \emph{list}, we instead
have a support \emph{stream}, which is its infinite.
\begin{romdefinition}[Stream]
  \begin{equation}
    \mathbf{Stream}(A) = \llbracket \top \triangleright \text{const} (\mathbb{N}) \rrbracket(A) \simeq \mathbb{N} \rightarrow A
  \end{equation}
\end{romdefinition}
\begin{romdefinition}[Split Countability]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \mathbf{Stream}(A))} , \Pi {(x : A)} , x \in \mathit{xs}
  \end{equation}
\end{romdefinition}
This type is definitionally equal to it surjection equivalent (\(\mathbb{N}
\twoheadrightarrow ! \; A\)).
\subsubsection{Closure}
We know that countable infinity is not closed under the exponential (function
arrow), so the only closure we need to prove is \(\Sigma\) to cover all of
what's left.
\input{figures/pairing-functions.tex}
\begin{romtheorem} \label{split-countability-sigma}
  Split countability is closed under \(\Sigma\).
\end{romtheorem}
\begin{proof}
  The main task here is to figure out a pattern which will pair elements of the
  two support streams without diverging.
  Figure~\ref{cartesian} illustrates why the pairing function we used for lists
  won't work: it needs to exhaust one of the input lists before inspecting
  anything other than the first element of the other.
  Of course, we can't exhaust a stream, so this diverges.

  Instead, we will use the pattern in figure~\ref{cantor}.
  We will produce a stream of lists (where each list is a diagonal), which is
  then concatenated to the final output stream.
  The intermediate stream is sometimes called the discrete convolution of the
  two input streams.
\end{proof}
\subsubsection{Kleene Star}
While we lose some closures with the inclusion of infinite types, we gain some
others.
In particular, we have the Kleene star.
This means that we have closure under lists.
\begin{romtheorem}
  Split countability is closed under Kleene star.
  \begin{equation}
    \mathcal{E}!(A) \rightarrow \mathcal{E}!(\mathbf{List}(A))
  \end{equation}
\end{romtheorem}
\begin{proof}
  As with the proof of closure under \(\Sigma\), our main task here is to figure
  out a pattern that reaches any given finite list in finite time.
  Our first step is to split up the problem as we did for \(\Sigma\)-closure: we
  produce an intermediate stream of lists, which we then concatenate to give the
  final result.
  The intermediate stream is constructed like so:
  \begin{equation}
    \mathit{xs}^\star_i = \left\{ \text{map}(\mathit{xs} , x) \mid \mathit{x} \in \mathbf{List}(\mathbb{N}) , \text{sum}(\text{map}(1+, x)) \equiv i \right\}
  \end{equation}
  In other words, the \(i\)th element is all lists such that the sum of the
  successor of the indices in the list is equal to \(i\).
\end{proof}
\subsection{Other Types of Countability}
There are of course other forms of countability available to us: manifest
countability, cardinal, etc.
We have a similar structure of relationships to the finite types, with the
exception that we cannot construct an isomorphism with \(\mathbb{N}\) fiven a
surjection from \(\mathbb{N}\).
\begin{itemize}
  \item Definition
  \item \(\Sigma\)
  \item Kleene star
  \item Truncated
  \item Topos
\end{itemize}
\section{Search}
\subsection{Omniscience}
In this section we are interested in restricted forms of the limited principle
of omniscience \cite{myhillErrettBishopFoundations1972}.
\begin{romdefinition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience is as follows:
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  \end{equation}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{romdefinition}
The limited principle of omniscience is non-constructive, but individual types
can themselves satisfy omniscience.
In particular, \emph{finite} types are omniscient.

There is also a universal form of omniscience, which we call exhaustibility.
\begin{romdefinition}[Exhaustibility]
  We say a type \(A\) is exhaustible if, for any decidable predicate \(P\) on
  \(A\), the universal quantification of the predicate is decidable.
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Pi {(x : A)} , P(x) \right)
  \end{equation}
\end{romdefinition}

All of the finiteness predicates we have seen justify exhaustibility.
We will only prove it once, then, for the weakest:
\begin{romtheorem}
  Kuratowski-finite types are exhaustible.
\end{romtheorem}

Omniscience is stronger than exhaustibility, as we can derive the latter from
the former:
\begin{romlemma} \label{omniscient-is-exhaustible}
  Any omniscient type is exhaustible.
\end{romlemma}
\begin{proof}
  For decidable propositions, we know the following:
  \begin{equation} \label{lem-conversion}
    \Pi (x : A) , P(x) \leftrightarrow \neg \Sigma (x : A) , \neg P(x) 
  \end{equation}
  To derive exhaustibility from omniscience, then, we run the predicate in its
  negated form, and then subsequently negate the result.
  The resulting decision over \(\neg \Sigma(x : A) , \neg P(x)\) can be
  converted into \(\Pi (x : A) , P(x)\).
\end{proof}
We cannot derive, however, that any exhaustible type is omniscient, as we do not
have the inverse of equation~\ref{lem-conversion}:
\begin{equation} \label{choice-conversion}
  \Sigma (x : A) , P(x) \leftrightarrow \neg \Pi (x : A) , \neg P(x)
\end{equation}
Such an equation would allow us to pick a representative element from any type,
which is therefore non-constructive.
In a sense, equation~\ref{lem-conversion} requires a form of LEM on the
proposition (i.e. requires it to be decidable), whereas
equation~\ref{choice-conversion} requires a form of choice.
Those finiteness predicates which are ordered do in fact give us this form of
choice, so the conversion is valid.
As such, all of the ordered finiteness predicates imply omniscience.
Again, we will prove it only for the weakest.
\begin{romtheorem}
  Manifest enumerable types are omniscient.
\end{romtheorem}
Finally, we do have a form of omniscience for prop-valued predicates, as they do
not care about the chosen representative.
\begin{romtheorem}
  Kuratowski finite types are omniscient about prop-valued predicates.
\end{romtheorem}
\subsection{Synthesising Pattern-Matching Proofs}
In particular, they can automate large proofs by analysing every possible case.
In \cite{firsovDependentlyTypedProgramming2015}, the \(\AgdaDatatype{Pauli}\)
group is used as an example.
\ExecuteMetaData[agda/Data/Pauli.tex]{def}
As \(\AgdaDatatype{Pauli}\) has 4 constructors, \(n\)-ary functions on
\(\AgdaDatatype{Pauli}\) may require up to \(4^n\) cases, making even simple
proofs prohibitively verbose.

The alternative is to derive the things we need from \(\mathcal{E}!\) somehow.
As \(\AgdaDatatype{Pauli}\) is a simple finite type, the instance can be defined
in a similar way to those in lemma~\ref{split-enum-bool-top-bot}.
From here we can already derive decidable equality, a function which requires 16
cases if implemented manually.

For proof search, the procedure is a well-known one in Agda
\cite{devrieseBrightSideType2011}: we ask for the result of a decision procedure
as an \emph{instance argument}, which will demand computation during
typechecking.
\subsection{Multiple Arguments}
The automation machinery above only deals with single-argument predicates.
This is not a problem, as we know that we can work with multiple arguments by
currying and uncurrying, since all of the finiteness predicates are closed under
\(\times\).
To automate away the curry/uncurry noise we will use instance search, building
on \cite{allaisGenericLevelPolymorphic2019} to develop a small interface to
generic \(n\)-ary functions and properties.
Our generic representation can handle dependent \(\Sigma\) and \(\prod\)
types (rather than their non-dependent counterparts, \(\times\) and
\(\rightarrow\)).
This extension was necessary for our use case: it is mentioned in the paper
as the obvious next step. 
We also implement the curry-uncurry combinators as (verified) isomorphisms.

A full explanation of our implementation is beyond the scope of this work, so we
only present the finished interface, which is used like so:
\ExecuteMetaData[agda/Data/Pauli.tex]{assoc-prf}
\subsection{Synthesising Functions}
Finally, thanks to extensionality provided to us by HoTT, and the computation
properties provided by CuTT, we can derive decidable equality on functions over
finite types.
We can also use functions in our proof search.
Here, for instance, is a automated procedure which finds the
\(\AgdaFunction{not}\) function on \(\mathbf{Bool}\), given a specification.
\ExecuteMetaData[agda/Data/Pauli.tex]{not-spec}
